Supporting Information : Complex and informative dependencies propogate throughout ecology and can be revealed using copulas

### 2019-05-13 Paper.Rmd
green spruce aphid-winter temp copula figures, table added

<!--***DAN: green spruce aphid all the way from causes to consequences-->
Our hypotheses cover propagation of copula structure from causes, 
to observable data, to consequences for human concerns.
We here take a closer look at the green spruce aphid, which illustrates 
all these stages in one example.
Green spruce aphid abundance, as detected in the suction trap data we use (Data,
Appendix \ref{SM-Data_details}),
is known to be strongly influenced by temperature the previous winter (@Sheppard2016,
supplementary fig. 6).
For each of our 10 sampling locations, we examined (Appendix \ref{SM-sect:greenspruce}, 
Fig. \ref{}) the copula of winter temperature
and green spruce aphid abundance time series for the location, finding 
substantial left-tail dependence: apparently winter temperature has an asymmetric
influence on abundance in the sense that cold winters produce low abundances
but warm winters do not yield especially higher abundances than moderate winters
(in passing, this is evidence for the first two parts of causal mechanism A).

>>>>

<!--***DAN: green spruce aphid all the way from causes to consequences-->
Our hypotheses cover propagation of copula structure from causes, 
to observable data, to consequences for human concerns.
We here take a closer look at the green spruce aphid, which illustrates 
all these stages in one example.
Green spruce aphid abundance, as detected in the suction trap data we use (Data,
Appendix \ref{SM-Data_details}),
is known to be strongly influenced by temperature the previous winter (@Sheppard2016,
supplementary fig. 6).
For each of our 10 sampling locations, we examined (Appendix \ref{SM-sect:greenspruce}, 
Fig. \ref{SM-fig:greenspruce}) the copula of winter temperature
and green spruce aphid abundance time series for the location, finding 
substantial left-tail dependence: apparently winter temperature has an asymmetric
influence on abundance in the sense that cold winters produce low abundances
but warm winters do not yield especially higher abundances than moderate winters
(in passing, this is evidence for the first two parts of causal mechanism A).

### 2019-05-13 SupportingInformation.Rmd
added another request for Shya, for a table on the green spruce aphid
green spruce aphid-winter temp copula figures, table added

# Aphid raw data for flight duration   
#d0_flightduration<-as.matrix(read.csv("./Data/Aphid_data/APHID_DATA_RAW/aphidtsflightduration141117.csv",header=F))

# Time series analysis aphids winter avg temp (here taken as -ve)
temp_aphid_raw<-as.matrix(read.csv("./Data/Aphid_data/APHID_DATA_RAW/aphidmonthlytempdata.csv",header = F))

@@

aphid_winter_neg_temp<-vector("list",11) #initializing
names(aphid_winter_neg_temp)<-rownames(winter_avg_temp)

for(i in 1:11){
  aphid_winter_neg_temp[[i]]<-data.frame(Year=c(1976:2010),Dat=-winter_avg_temp[i,])
}

temp_aphid<-list(aphid_winter_neg_temp)
names(temp_aphid)<-"aphid_winter_neg_temp"
saveRDS(temp_aphid,"./Results/temp_aphid_winter_neg_temp_avg_data.RDS")
```

@@

# From causes to consequences for green spruce aphid abundance data\label{sect:greenspruce}
Describe the winter temp data
Show the copulas with winter temp, 10 locations

<!--***DAN: Shya, please insert here a figure showing the 10 copulas between winter
temp and green aphid count data as 10 panels. Insert the code that generates that figure 
in an appropriate place so it is part of the overall workflow. Please use the label 
fig:greenspruce, since that is what I refer to this fig by in the main text.-->

>>>>

# Time series analysis aphids winter avg temp 
temp_aphid_raw<-as.matrix(read.csv("./Data/Aphid_data/APHID_DATA_RAW/aphidmonthlytempdata.csv",header = F))

@@

aphid_winter_temp<-vector("list",11) #initializing
names(aphid_winter_temp)<-rownames(winter_avg_temp)

for(i in 1:11){
  aphid_winter_temp[[i]]<-data.frame(Year=c(1976:2010),Dat=winter_avg_temp[i,])
}

temp_aphid<-list(aphid_winter_temp)

#-------------------------------------------------------------------------
# append winter avg temp data as sp 21 in aphid's count data

data_aphid_count_with_temp<-append(data_aphid_count,temp_aphid)
names(data_aphid_count_with_temp)[[21]]<-"winter_avg_temp"

# create res folder
if(!dir.exists("./Results/aphid_count_winter_temp_results")){
  dir.create("./Results/aphid_count_winter_temp_results/")
}

saveRDS(data_aphid_count_with_temp,"./Results/aphid_count_winter_temp_results/data_aphid_count_with_temp.RDS")
```

@@

# From causes to consequences for green spruce aphid abundance data\label{sect:greenspruce}

For each locations (selected based on a threshold of 30 years of data, see \ref{Data_details}), 
we computed yearwise average temperature of winter season (spanning over 4 months from December, previous 
year to March, present year). With this winter temperature time series and green spruce aphid abundance time series data 
a bivariate copula was made for each selected locations and shown in Fig. \ref{fig:greenspruce} whenever it was 
significantly dependent with a p-value $<0.05$. Stronger lower tail dependence
was visually prominent in those copulas. We used our model selection and non-parametric statistical tools 
(see Table \ref{tab:greenspruce}) to confirm our hypothesis that winter temperature 
was the main driver in "lower tail dependence" of the green
spruce aphid abundance copula from any two location pairs (see Table \ref{tab_count_LTmUT}).

```{r copulaplot_greenspruceaphid_with_temp,echo=F,results="hide"}
source("./getcopula.R")
source("./good_loclist.R")
sp<-10 #only for green spruce aphid

aphid_temp_coplist<-vector("list",1) 
names(aphid_temp_coplist)<-c("sp10_Green_spruce_aphid")

loclist<-good_loclist(d_allsp=data_aphid_count_with_temp,sp=sp,data_pt_thrs=30)

pdf("./Results/aphid_count_winter_temp_results/copulaplot_greenspruceaphidcount_temp.pdf",height=4,width = 10)
op<-par(mfrow=c(2,5),mar=c(3,3,3,3), mgp=c(1.5,0.5,0)) # each row for each sp for 2:11 locations

  gsa_temp_coplist<-vector("list",10) # for 10 selective locations which had atleast 30 datapoints
  names(gsa_temp_coplist)<-paste("loc",loclist,sep="")

  for(iloc in c(1:length(loclist))){
    loc<-loclist[iloc]
    s1<-cbind(data_aphid_count_with_temp[[sp]][[loc]]$Dat,data_aphid_count_with_temp[[21]][[loc]]$Dat)
    c1<-getcopula(d=s1,rankon=T,ploton=F)
    vi<-c1[,1]
    vj<-c1[,2]

    IndepTestRes<-VineCopula::BiCopIndTest(vi,vj)$p.value
    ct<-cor.test(vi,vj,alternative = "two.sided",method="kendall",exact=F)
    tauval<-unname(ct$estimate)

      if(IndepTestRes<0.05 && tauval>0){
        plot(vi,vj,type='p',col=rgb(0,0,0,0.3),pch=19,xlim=c(0,1),ylim=c(0,1),
             xlab=names(data_aphid_count_with_temp)[sp],ylab="winter temp.",cex.lab=1.5)
        mtext(paste0("Loc = ",loc," , p = ",round(IndepTestRes,3)),side = 3, line=0.15, adj=0.5)
      }else if(IndepTestRes<0.05 && tauval<0){
        plot(vi,vj,type='p',col=rgb(0,1,0,0.3),pch=19,xlim=c(0,1),ylim=c(0,1),
             xlab=names(data_aphid_count_with_temp)[sp],ylab="winter temp.",cex.lab=1.5)
        mtext(paste0("Location = ",loc," , p = ",round(IndepTestRes,3)),side = 3, line=0.15, adj=0.5)
      }else{
        plot(-1,0,xlim=c(0,1),ylim=c(0,1),xlab=names(data_aphid_count_with_temp)[sp],ylab="winter temp.",cex.lab=1.5)
        text(0.5,0.5,"Indep.",adj=c(0.5,.5),cex=2)
        mtext(paste0("Location = ",loc," , p = ",round(IndepTestRes,3)),side = 3, line=0.15, adj=0.5)
      }

    gsa_temp_coplist[[iloc]]<-c1
  }
aphid_temp_coplist[[1]]<- gsa_temp_coplist
par(op)
dev.off()
```

\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm]{./Results/aphid_count_winter_temp_results/copulaplot_greenspruceaphidcount_temp.pdf}
\caption[Copula plot for green spruce aphid abundance data with winter temperature]{Copula plot for green spruce 
aphid abundance data with winter temperature for each selected locations. \label{fig:greenspruce}}
\end{center}
\end{figure}

```{r ms_npa_greenspruceaphid_count_temp, echo=F, results="hide", cache=T, cache.extra=list(seed,aphid_temp_coplist,mtime("CopulaFunctions_flexible.R"),mtime("OurBiCopSelect.R"),mtime("MyBiCopGofTest.R"),mtime("preprocessors.R"))}

set.seed(seed)
source("./OurBiCopSelect.R")
source("./CopulaFunctions_flexible.R")

#====================== FUNCTION TO GET MS RESULTS FOR Green spruce aphid abundance with temp copula from all 10 locs===========================

get_ms_aphid_temp<-function(sp,d_copsp){

  cop_eachsp_allloc<-d_copsp[[sp]]

  lengoodloc<-length(cop_eachsp_allloc)

  # model selection with a copula (made of aphid-count and winter avg temp ts) from each location 2:11
  ms_summary_allloc<-c()
  ms_res_allloc<-vector("list",lengoodloc)
  names(ms_res_allloc)<-paste("Loc",c(2:11),sep="")

  for(loc in 1:lengoodloc){

    cop_eachloc<-cop_eachsp_allloc[[loc]]
    u1<-cop_eachloc[,1]
    u2<-cop_eachloc[,2]

    #----------------- model selection -----------------------------
    ms_res<-OurBiCopSelect(u1=u1,u2=u2,families=c(1,3:10,13:14,16:20),
                           level=0.05,AICBIC="AIC",numBSsmall=100,pthresh=0.2,numBSlarge=1000,
                           gofnormal=F,status=F)
    ms_res_allloc[[loc]]<-ms_res

    m<-as.data.frame(matrix(NA,1,12))
    colnames(m)<-c("p.Indep","minAIC.bestfit.cop","Bestfit.cop.AICw",
                   "Normal.cop.AICw","p.CvM","p.KS",
                   "LT.model.avg","UT.model.avg","(LT-UT).model.avg",
                   "CorlmCoru","PlmPu","D2umD2l")
    m$p.Indep<-ms_res$IndepTestRes

    if(ms_res$IndepTestRes<0.05){
      #------------------------------ non-param analysis ---------------------------
      #Cor stat
      Corl<-Corbds(vi=u1,vj=u2,lb=0,ub=0.5)
      Coru<-Corbds(vi=u1,vj=u2,lb=0.5,ub=1)
      CorlmCoru<-Corl-Coru

      #P stat
      res_l<-Pbds(vi=u1,vj=u2,lb=0,ub=0.5)
      Pl<-res_l$abs_res

      res_u<-Pbds(vi=u1,vj=u2,lb=0.5,ub=1)
      Pu<-res_u$abs_res
      PlmPu<-Pl-Pu

      #D2 stat
      D2l<-D2bds(vi=u1,vj=u2,lb=0,ub=0.5)
      D2u<-D2bds(vi=u1,vj=u2,lb=0.5,ub=1)
      D2umD2l<-D2u-D2l

      m$minAIC.bestfit.cop<-ms_res$InfCritRes$copname[which.min(ms_res$InfCritRes$AIC)]
      m$Bestfit.cop.AICw<-round(max(ms_res$InfCritRes$AICw),3)
      m$Normal.cop.AICw<-round(ms_res$InfCritRes$AICw[1],3)
      m$p.CvM<-ms_res$GofRes_CvM
      m$p.KS<-ms_res$GofRes_KS
      m$LT.model.avg<-round(ms_res$relLTdep_AICw,3)
      m$UT.model.avg<-round(ms_res$relUTdep_AICw,3)
      m$`(LT-UT).model.avg`<-m$LT.model.avg-m$UT.model.avg
      m$CorlmCoru<-CorlmCoru
      m$PlmPu<-PlmPu
      m$D2umD2l<-D2umD2l
    }

    ms_summary_allloc<-rbind(ms_summary_allloc,m)
    #cat("loc=",loc,"=========ok=========","\n")

  }
  rownames(ms_summary_allloc)<-names(ms_res_allloc)
  return(list(ms_res_allloc=ms_res_allloc,
              ms_summary_allloc=ms_summary_allloc))
}

#===================== call above function for green spruce aphid abundance and winter temp data =================

d_copsp<-aphid_temp_coplist
splist<-c(1:length(d_copsp))
ms_npa_RES_greenspruceaphid_count_temp<-vector("list",length(splist))
names(ms_npa_RES_greenspruceaphid_count_temp)<-c("sp10_Green_spruce_aphid")

for(isp in splist){
  #cat("--------------sp =",isp,"-----------------------","\n")
  res<-get_ms_aphid_temp(sp=isp,d_copsp=d_copsp)
  ms_npa_RES_greenspruceaphid_count_temp[[isp]]<-res
}

saveRDS(ms_npa_RES_greenspruceaphid_count_temp,"./Results/aphid_count_winter_temp_results/ms_npa_RES_greenspruceaphid_count_temp.RDS")
```

```{r tab_greenspruceaphid_temp_summary, echo=F, results='asis',message=F}
library(kableExtra)
set.seed(seed)

res<-ms_npa_RES_greenspruceaphid_count_temp$sp10_Green_spruce_aphid$ms_summary_allloc
#row.names(res)<-NULL
#res<-cbind(c(2:11),res)
colnames(res)<-c("p-value from indep. test","bestfit copula",
                 "bestfit copula AICw","Normal Copula AICw","p-value from CvM stats","p value from KS stats","LT dependence","UT dependence","(LT - UT) dependence","cor$_{l}$ - cor$_{u}$","P$_{l}$ - P$_{u}$","D$_{u}^2$ - D$_{l}^2$")
res<-res[,-1]
knitr::kable(res,"latex", booktabs = T, linesep = "",digits=3, escape=F,longtable=T,
             caption.short="Summary results for copula made with time-series of \\textbf{green spruce aphid} abundance data and winter temperature for each selected locations",
      caption = "Summary results for copula made with time-series of green spruce aphid abundance data and winter temperature for each selected locations, tail dependence were calculated using model averaged AIC-weights, NA represents copula was independent at given location as p-value from independence test was beyond the threshold 0.05. Last three columns indicate non-parametric statistics computed on $(0,0.5)$ and $(0.5,1)$ bounds for lower and upper statistics, respectively.\\label{tab:greenspruce}") %>% column_spec(1:12,width="1.2 cm")#%>%
#kable_styling(latex_options = c("scale_down"))
```
